Eco-evolutionary dynamics of finite populations from first principles

### 2023-02-04 ind_files/03_BD_processes.tex
Separate field eqns into separate chapter

Raymond Pearl, one of the pioneers of mathematical ecology as a discipline, wrote in 1927 that ``What we want to know is how the biological forces of natality and mortality are so integrated and correlated in their action as to lead to a final result in size of population which follows this particular curve rather than some other one'' \citep{pearl_growth_1976}. Pearl realized 95 years ago that population dynamics must be ultimately explained by the mechanistic processes of birth and death. Today, there are mounting calls for more mechanistic models of evolution that are grounded in these fundamental processes of birth and death \citep{geritz_mathematical_2012,doebeli_towards_2017}. Ecologists also increasingly recognize that incorporating stochasticity is vital to developing more realistic ecological models \citep{hastings_transients_2004, coulson_skeletons_2004, boettiger_noise_2018, shoemaker_integrating_2020,schreiber_does_2022} and does more than `add noise' to deterministic expectations. Individual-based models, where ecological rules are specified at the level of the individual, are a powerful mathematical tool for mechanistic descriptions of stochastic population dynamics \citep{black_stochastic_2012}. Birth-death processes are a very general class of stochastic processes that can be used to capture a wide range of eco-evolutionary processes. `System-size expansions' and their subsequent analysis using the `weak noise approximation' are common tools for analyzing stochastic birth-death processes that are well-known in the statistical physics and applied mathematics communities \citep{gardiner_stochastic_2009}. However, these tools remain relatively underappreciated in ecology, despite being relatively easy to understand and extremely well-motivated in scenarios germane to ecology and evolution. Here, I present a formulation of population dynamics constructed from first principles grounded in birth-death processes. To facilitate readership by a broad audience, only passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability are assumed. Familiarity with stochastic calculus is helpful for some sections but is not required, and I provide a brief introduction below. I begin by introducing birth-death processes, SDEs, and the Fokker-Planck equation in a non-technical manner, and providing the intuition for system-size expansions and weak noise analysis in ecological systems. In section \ref{sec_1D_processes}, I show how fluctuating population size of populations of identical individuals can be tracked through a one-dimensional birth-death process. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process. As an example, I illustrate the complete process for a stochastic version of the logistic equation. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, illustrate the system size expansion and the weak noise approximation. Under mild assumptions, we show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, Eigen's quasispecies equation), thus establishing the microscopic basis of well-known equations in evolutionary game theory. I also show that the mean value of the trait in the population changes according to the Price equation in the deterministic limit. Section \ref{sec_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck equation, in terms of functional derivatives. Under mild assumptions, we show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. We also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation and illustrate the process using the stochastic quantitative logistic equation introduced by \citep{doebeli_adaptive_2011} and \citep{rogers_demographic_2012} as an example.  Sections \ref{sec_1D_processes} and \ref{sec_nD_processes} provide a unified, pedagogical review and conceptual synthesis of techniques that are well-known in physics, as applied to population biology. Section \ref{sec_infD_processes}, to the best of my knowledge, is entirely original. Section \ref{sec_infD_processes} also presents a simple heuristic derivation of quantitative genetics models and adaptive dynamics from stochastic first principles that is simpler than the rigorous mathematical derivations grounded in measure theory and martingale/markov theory that are currently standard reference in theoretical ecology \citep{champagnat_individual_2008}.

@@

\section{One-dimensional processes for population size}\label{sec_1D_processes}
The simplest birth-death processes are those in which the state at any time can be characterized by a single number. Populations of identical individuals are an obvious example of such a system. The mathematics below are adapted from sections 6.3 and 7.2 of \citep{gardiner_stochastic_2009}.

@@

\section{Infinite-dimensional processes for quantitative traits}\label{sec_infD_processes}
Things become more complicated when we deal with `quantitative' traits. Traits like body size, body weight, or beak length, often take on uncountably many values (say, all values in the interval $[0,1]$, for example). In this case, we cannot describe the population using a vector but instead require a function. More precisely, if the set of all possible trait values is $\mathcal{T}$, we will characterize the population using a special kind of function $\phi^{(t)}$ such that the quantity $\int_{A}\phi^{(t)}(x)dx$ gives us the number of individuals that are in any `nice' region $A \subset \mathcal{T}$ of the possible trait space\footnote{The mathematically informed reader may notice that this sounds like I am trying to dance around the word `measure'. Indeed, we are really looking to construct branching processes that take values in some nice space of measures that can be endowed with 
sufficient mathematical structure for notions like convergence and integration to make sense. All the Dirac deltas that will turn up shortly are `properly' viewed as measures, and integrals with Dirac deltas in the integrand are to be interpreted as integration with respect to the Dirac measure. If one tries to be careful about these things, they will quickly find themselves drowning in a quagmire of mathematical formalism.
% For example, one can't really justify using informal tools like the functional derivative, which is ill-defined and `properly' interpreted as a Fr\'echet derivative in a suitable normed function space (making the use in conjunction with Dirac measures/distributions somewhat dubious without further elaboration on the space and norm in question). The existence of a probability density function for this process is also \emph{a priori} somewhat questionable and requires detailed arguments using tools like the Radon-Nikodym theorem. Here, we adopt the physicist's solution of ignoring all such potential difficulties and assuming a clever mathematician can make sure everything works out using arcane tools such as ``functional analysis" and  ``distribution theory".
If you know and care about enough mathematics for this to really bother you, see \citep{champagnat_unifying_2006} for a much more rigorous treatment that avoids using informal tools such as functional derivatives and functional equivalents of Fokker-Planck equations in favor of a probabilistic approach grounded in (measure-theoretic) Markov and martingale theory.}. The state space of the stochastic process thus becomes infinite-dimensional, which complicates matters slightly. The principal objects of interest here are \emph{functionals} $F[x, \phi^{(t)}]$ which take in a scalar $x$ representing the trait value of interest, and a function $\phi^{(t)}$ representing the population at time $t$. Thus, whereas in the previous section we were interested in how a function $f(x(t))$ changes based on the change in an input variable $x(t)$ (the population), we are now interested in how a functional $F[\phi^{(t)}]$ changes with the change in an input function $\phi^{(t)}$. The appropriate tool for this notion is the functional derivative $\delta F/\delta \phi$. The functional derivative is an \emph{ad hoc}, somewhat informal notion, defined indirectly as the unique object that obeys, for any `nice' function $\rho$
\begin{equation}
\label{functional_derivative_defn}
    \int\frac{\delta F}{\delta \phi(x)}\rho(x)dx = \lim_{h \to 0} \frac{F[\phi + h\rho]-F[\phi]}{h}
\end{equation}
This definition is formulated in analogy to directional derivatives in multi-variable calculus: Noting that a function can be thought of as an infinite-dimensional vector, informally `taking the limit' $n \to \infty$ in \eqref{directional_derivative_defn} yields \eqref{functional_derivative_defn}.
We envision a population of individuals with a `trait' that takes values in some one-dimensional set $\mathcal{T} \subseteq \mathbb{R}$. Since the trait of any given individual is fixed, and since each individual can only have one exact trait value, an individual with a trait value $x \in \mathcal{T}$ can be characterized as a Dirac delta mass centered at $x$, defined indirectly as the object which satisfies, for any one-dimensional function $f$,
    \int\limits_{A}f(y)\delta_{x_i}dy = 
    \begin{cases}
        f(x_i) & x_i \in A\\
        0 & x_i \notin A 
    \end{cases}
for every `nice' subset $A \subset \mathcal{T}$. In physics notation, we would write $\delta_{x_i} = \delta(y-x_i)$ as a `function' of the dummy variable $y$ (which will be integrated over). Note that by choosing $f(x) \equiv 1$, we get an `indicator' that is $1$ if the individual is within the set $A$ and $0$ otherwise. Thus, if the population at any time $t$ consists of $N(t)$ individuals with trait values $\{x_1,x_2,\ldots,x_{N(t)}\}$, then it can be completely characterized (Figure \ref{fig_infD_pop_description}) by the `distribution'
    \nu^{(t)} = \sum\limits_{i=1}^{N(t)}\delta_{x_i}
which in physics notation would be a function\footnote{Physicists use the term `field' for functions of the form $f(x,t):\mathbb{R}^{n} \times [0,\infty) \to \mathbb{R}^{m}$, where $\mathbb{R}^{n}$ represents space and $[0,\infty)$ represents time. They then call models which describe such functions `field theories'. In physics jargon, our stochastic process $\{\nu^{(t)}\}_{t \geq 0}$ when viewed as a sequence of functions $\{\nu^{(t)}(y)\}_{t \geq 0}$ thus describes a (scalar) `stochastic field', and the formalism we will develop below is a `stochastic field theory' of evolution, where physical space has been replaced by an abstract trait space. This is closely related to the area of physics called `statistical field theory', the analog of quantum field theory for systems with a large number of classical particles. Stochastic field theories over physical space have recently been used in biology to model brain function \citep{bressloff_stochastic_2010} and collective motion \citep{o_laighleis_minimal_2018}.} $\nu^{(t)}(y) = \sum_{i}^{N(t)} \delta(y-x_i)$. Thus, the state space of our process is
    \mathcal{M} = \left\{\sum\limits_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}
Note that for any set $A \subset \mathcal{T}$, $\int_A\nu^{(t)}dx$ gives the number of individuals that have trait values that lie within the set $A$ and that integrating over $\mathcal{T}$ gives the population size $N(t)$ at time $t$. Given the population $\nu^{(t)} = \sum_{i=1}^{N(t)}\delta_{x_i}$ and a real function $f(x)$, we have $\int_{\mathcal{T}}f(y)\nu^{(t)}dy = \sum_{i=1}^{N(t)}f(x_i)$.
\myfig{0.5}{Media/3.3_dirac_deltas.png}{\textbf{Schematic description of a function valued birth-death process.} Consider a population of birds in which individuals have varying beak lengths. \textbf{(A)} Each individual in the population can be described as a Dirac delta mass centered at its beak length. This is because each individual has exactly one fixed beak length, and therefore, can be thought of as a distribution centered at that particular beak length and with zero spread. \textbf{(B)} The population as a whole is thus described as a sum of Dirac masses. $N(t)$ here is the size of the population at time $t$. Birth and death of individuals would correspond to the addition and removal of Dirac masses respectively. Note that if we had a large number of individuals, this distribution begins to look like a continuous distribution.}{fig_infD_pop_description}
% We are interested in observing the behaviour of elements of
%     \mathcal{F} = \left\{f:\mathcal{M} \to \mathbb{R} \ | \ f \ \textrm{bounded and measurable}\right\},
% the set of all bounded measurable real functions on $\mathcal{M}$. Given the population $\nu^{(t)} = \sum_{i=1}^{N(t)}\delta_{x_i}$ and a function $f \in \mathcal{F}$, we have $\int_{\mathcal{T}}f(x)\nu^{(t)}(dx) = \sum_{i=1}^{N(t)}f(x_i)$. Elements of $\mathcal{F}$ are interesting because they yield various quantities of biological relevance when integrated over $\mathcal{T}$ with respect to our population measure $\nu$. For example, the function $f(\nu^{(t)}) = 1$ gives the population size at time $t$, and the function $f(\nu^{(t)}) = \chi_{A}/N(t)$, where $\chi$ is the indicator function, gives the fraction of the population that have trait values in $A$.\\
Now that we have described the population, we must define the rules for how it changes. We assume that there exist non-negative functionals $b(x|\nu)$ and $d(x|\nu)$ which describe the rate at which individuals with trait value $x$ are born and die respectively in a population $\nu$. Again, we must be careful about what exactly we mean when we speak about `rates'. In this case, we mean that if we know that the population is currently described by the function $\nu$, and we know that \emph{either a birth or a death} occurs, then the probability that this event is the birth of an individual whose phenotype is within the set $A \subset \mathcal{T}$ is given by
    \mathbb{P}[\textrm{ Birth with offspring in }A | \textrm{ something happened }] = \frac{1}{\mathcal{N}}\int\limits_{A}b(x|\nu)dx
and the probability that the event is the death of an individual whose phenotype is within the set $A$ is
    \mathbb{P}[\textrm{ Death of an individual in }A | \textrm{ something happened }] = \frac{1}{\mathcal{N}}\int\limits_{A}d(x|\nu)dx
where $\mathcal{N} = \int_{\mathcal{T}}b(x|\nu)+d(x|\nu)dx$ is the normalizing constant in both cases. Note that we assume $\mathcal{N}$ is always finite and non-zero.

\begin{example}
Consider the birth and death functionals:
\begin{equation}
\label{Rogers_logistic_BD}
\begin{aligned}
b(x|\nu) &= r\int\limits_{\mathcal{T}}m(x,y)\nu(y)dy; \ m(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{m}^{2}}\right)\\
    d(x|\nu) &= \frac{\nu(x)}{Kn(x)}\int\limits_{\mathcal{T}}\alpha(x,y)\nu(y)dy; \ \alpha(x,y) = \exp\left(\frac{-(x-y)^2}{\sigma_{\alpha}^{2}}\right)
\end{aligned}
\end{equation}
This choice corresponds to an asexual population having a constant (per-capita) birth rate $r$. Birth is sometimes with mutation, and the extent of the mutations is controlled by a Gaussian kernel $m(x,y)$. The death rate is density-dependent, mediated by a Gaussian competition kernel $\alpha(x,y)$, and also contains a phenotype-dependent carrying capacity controlled by $n(x)$, scaled by a constant $K$. The biological interpretation of the death rate is through ecological specialization for limiting resources - Individuals have different intrinsic advantages (controlled by $n(x)$), and experience greater competition from conspecifics that are closer to them in phenotype space (controlled by $\alpha(x,y)$).
\end{example}
Let us now define, for each $x \in \mathcal{T}$, two \emph{step operators} $\mathcal{E}_{x}^{\pm}$ that satisfy
\begin{equation*}
    \mathcal{E}_{x}^{\pm}[f(y,\nu)] =  f(y,\nu \pm \delta_x)
\end{equation*}
In other words, the step operators $\mathcal{E}_{x}^{\pm}$ simply describe the effect of adding or removing a single individual with trait value $x$ from the population.
It is known (only for one-dimensional traits) that we can find a density function $P(\nu,t)$ such that the probability that the process takes value $\nu^{(t)}$ at time $t$ is given by $\int_{\mathcal{T}}P(\nu,t)dx$.\\
We can now use the same trick as earlier and obtain a master equation by counting inflow and outflow of states. Any change to a state must be through the addition or subtraction of a single Dirac delta mass. For any state $\nu \in \mathcal{M}$, the transition rate from $\nu - \delta_{x}$ to $\nu$ is simply $\mathcal{E}^{-}_{x}b(x|\nu)$, and similarly, the transition rate from $\nu+\delta_{x}$ to $\nu$ is $\mathcal{E}^{+}_{x}d(x|\nu)$. The transition rate out of $\nu$ to a state $\nu+\delta_x$ is just $b(x|\nu)$, and transition out to a state $\nu - \delta_x$ is just $d(x|\nu)$. Thus, integrating over all possible $x$ to obtain the total inflow and outflow rate for a state $\nu$, we see that $P(\nu,t)$ must satisfy:
\begin{equation}
\label{unnormalized_M_equation}
    \frac{\partial P}{\partial t}(\nu,t) = \int\limits_{\mathcal{T}}\left[(\mathcal{E}^{-}_{x}-1)b(x|\nu)P(\nu,t) + (\mathcal{E}^{+}_{x}-1)d(x|\nu)P(\nu,t)\right]dx
\end{equation}
This is the `Master equation' of our process.\\
\\
\subsection{The functional system-size expansion}

To proceed, as before, we assume that there exists a system-size parameter $K > 0$ to obtain a new process $\{\phi^{(t)}\}_{t \geq 0}$ such that for any set $A \subset \mathcal{T}$, $\int_A\phi^{(t)}dx$ gives the `density' of individuals that have trait values that lie within the set $A$. Note that we expect this stochastic process to evolve continuously if $K$ is large since the contribution of each individual is negligible. Specifically, we assume that there exists a $K>0$ such that we can make the substitutions:
\begin{align*}
    \phi^{(t)} = \frac{1}{K}\nu^{(t)} &= \frac{1}{K}\sum\limits_{i=1}^{N(t)}\delta_{x_i}\\
    b_K(x|\phi^{(t)}) &= \frac{1}{K}b(x|\nu^{(t)})\\
    d_K(x|\phi^{(t)}) &= \frac{1}{K}d(x|\nu^{(t)})
\end{align*}
$\{\phi^{(t)}\}_{t\geq0}$ takes values in 
\begin{equation*}
    \mathcal{M}_K = \left\{\frac{1}{K}\sum\limits_{i=1}^{n}\delta_{x_i} \ | \ n \in \mathbb{N}, x_i \in \mathcal{T}\right\}
\end{equation*}
In terms of these new variables, we obtain the master equation:
\begin{equation}
\label{M_equation}
    \frac{\partial P}{\partial t}(\phi,t) = K\int\limits_{\mathcal{T}}\left[(\Delta^{-}_{x}-1)b_K(x|\phi)P(\phi,t) +(\Delta^{+}_{x}-1)d_K(x|\phi)P(\phi,t)\right]dx
\end{equation}
where we have introduced new step operators $\Delta_{x}^{\pm}$ that satisfy:
\begin{equation*}
    \Delta_{x}^{\pm}[F(y,\phi)] =  F\left(y,\phi \pm \frac{1}{K}\delta_x\right)
\end{equation*}
\\
\\
We can now conduct a system-size expansion as before by using a functional `Taylor expansion' of the step operators. Recall that the functional version of the Taylor expansion of a functional $F[\rho]$ about a function $\rho_0$ defined on a domain $\Omega \subseteq \mathbb{R}$ is given by:
\begin{equation*}
    F[\rho_0 + \rho] = F[\rho_0] + \int\limits_{\Omega}\rho(x)\frac{\delta F}{\delta \rho_0(x)}dx + \frac{1}{2!}\int\limits_{\Omega}\int\limits_{\Omega}\rho(x)\rho(y)\frac{\delta^2 F}{\delta \rho_0(x)\delta \rho_0(y)}dxdy + \cdots
\end{equation*}
Since $\Delta^{\pm}_{x}[F[\phi]] = F[\phi \pm \delta_x/K]$, we can Taylor expand the RHS to see that our step operators obey
\begin{align}
    \Delta^{\pm}_{x}[F[\phi]] &= F[\phi] \pm \frac{1}{K}\int\limits_{\mathcal{T}}\frac{\delta F}{\delta \phi(y)}\delta_xdy + \frac{1}{2K^2}\int\limits_{\mathcal{T}}\int\limits_{\mathcal{T}}\frac{\delta^2 F}{\delta \phi(y)\delta \phi(z)}\delta_xdy\delta_xdz+\mathcal{O}(K^{-3})\nonumber\\
    &= F[\phi] \pm \frac{1}{K}\frac{\delta F}{\delta \phi(x)} + \frac{1}{2K^2}\frac{\delta^2 F}{\delta \phi(x)^2}+\mathcal{O}(K^{-3})
    \label{KM_ansatz}
\end{align}
Neglecting terms of $\mathcal{O}(K^{-3})$, we can now substitute \eqref{KM_ansatz} into \eqref{M_equation} to obtain:
\begin{equation*}
\begin{split}
\frac{\partial P}{\partial t}(\phi,t) = K\int\limits_{\mathcal{T}}\left[
    \left(-\frac{1}{K}\frac{\delta}{\delta\phi(x)} + \frac{1}{2K^2}\frac{\delta^2}{\delta\phi(x)^2}\right)\{b_K(x|\phi)P(\phi,t)\}\right]dx\\
    +K\int\limits_{\mathcal{T}}\left[\left(\frac{1}{K}\frac{\delta}{\delta\phi(x)} + \frac{1}{2K^2}\frac{\delta^2}{\delta\phi^2(x)}\right)\{d_K(x|\phi)P(\phi,t)\}\right]dx
\end{split}
\end{equation*}
Rearranging these terms, we obtain a `functional Fokker-Planck equation':
\begin{equation}
\label{functional_FPE}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial P}{\partial t}(\phi,t) = \int\limits_{\mathcal{T}}\left[-
    \frac{\delta}{\delta\phi(x)}\{\mathcal{A}^{-}(x|\phi)P(\phi,t)\} + \frac{1}{2K}\frac{\delta^2}{\delta\phi(x)^2}\{\mathcal{A}^{+}(x|\phi)P(\phi,t)\}\right]dx}
\end{equation}
where
\begin{align*}
   \mathcal{A}^{\pm}(x|\phi) &= b_K(x|\phi)\pm d_K(x|\phi) = \frac{1}{K}\left(b(x|\nu)\pm d(x|\nu)\right)
\end{align*}

This constitutes the `mesoscopic' description. For large (but finite) $K$, equation \eqref{functional_FPE} can be analyzed using a weak noise approximation as before.

\subsection{The deterministic limit}
We can once again appeal to the link between Fokker-Planck equations and Langevins \citep{lafuerza_role_2016} to say that \eqref{functional_FPE} corresponds to the Langevin equation:
\begin{equation}
\label{functional_langevin}
    \frac{\partial \phi}{\partial t}(x,t) = \mathcal{A}^{-}(x|\phi) + \frac{1}{\sqrt{K}}\eta(x,t)
\end{equation}
where $\eta(x,t)$  is the `Gaussian spacetime white noise' with zero mean and autocovariance function
\begin{equation*}
    \mathbb{E}[\eta(x,t)\eta(x',t')] = \sqrt{\mathcal{A}^{+}(x|\phi)\mathcal{A}^{+}(x'|\phi)}\delta(x-x')\delta(t-t')
\end{equation*}
Taking $K \to \infty$ in equation \eqref{functional_langevin} then yields a PDE:
\begin{equation}
\label{deterministic_traj}
\frac{\partial \psi}{\partial t}(x,t) = \mathcal{A}^{-}\left(x|\psi\right) = b_K(x|\psi)- d_K(x|\psi)
\end{equation}
where we have used a different symbol $\psi$ simply to highlight that $\psi(x,t)$ as the solution to equation \eqref{deterministic_traj} is a deterministic function, whereas $\phi(x,t)$ as defined in equation \eqref{functional_langevin} is really a stochastic process $\{\phi^{(t)}\}_{t\geq0}$. Equation \eqref{deterministic_traj} simply says that in the absence of stochasticity, the change in the density of individuals with trait values $x$ is given by the difference between the birth and death rates of these individuals in the population. Models of this form are precisely the `oligomorphic dynamics' of \citep{sasaki_oligomorphic_2011} if one assumes the population is composed of a small number of `morphs', \emph{i.e.} $\psi(x,t) = \sum\limits_{k=1}^{S} n_{k}\psi_k(x,t)$, where $\psi_k(x,t)$ is the phenotypic distribution of the $k$th morph (often assumed a normal distribution with narrow variance) and $S$ is the number of distinct morphs in the population. Equation \eqref{deterministic_traj} also appears in models of intraspecific trait variation in community ecology, such as the `trait space equations' of \citep{wickman_theoretical_2022} in their framework for eco-evolutionary community dynamics.\\
If one wishes to be mathematically careful, the connection between \eqref{functional_FPE} and \eqref{functional_langevin} becomes somewhat tenuous. In particular, while the equivalence between Fokker-Planck equations and SDEs (Langevin equations) is well-studied for finite-dimensional stochastic processes, it is much less understood for the infinite-dimensional function-valued processes that we are dealing with, and the interpretation of any formal `Langevin equation' (corresponding now to a stochastic partial differential equation, or SPDE) that we write down is unclear\footnote{The essential difficulty is interpreting what spacetime white noise formally means. See \citep{week_white_2021} for analytical progress for eco-evolutionary models via weak solutions to SPDEs (the rigorous theory needs some knowledge of measure theory and stochastic processes, but Week \emph{et al.} provide some very useful heuristics that only need Riemann integration, probability, and related notions)}. Nevertheless, we will pretend all is well and assume that one can do this, bolstered by the fact that we can recover some well-known deterministic equations from equation \eqref{deterministic_traj}, as we show below.

\subsubsection{Some familiar faces: Kimura-Crow and adaptive dynamics}

The `macroscopic' version of our function-valued process corresponds to quantitative genetics and adaptive dynamics (which is a generalization of evolutionary game theory for quantitative traits). We begin with the deterministic process given by \eqref{deterministic_traj}. We assume that the birth and death functions take the form:
\label{BD_for_kimura}
b_K(x|\psi) &= \mu b_{\textrm{mut}}(x|\psi) + b_{\textrm{ind}}(x|\psi)\psi(x,t)\\
d_K(x|\psi) &= d_{\textrm{ind}}(x|\psi)\psi(x,t)
Here, $ b_{\textrm{mut}}(x|\psi)$ describes birth due to mutations, and $\mu \geq 0$ is a mutation rate. The functions $b_{\textrm{ind}}(x|\psi)$ and $d_{\textrm{ind}}(x|\psi)$ describe the per-capita birth rate and death rate of type $x$ in a population $\psi$. For example,  $b_{\textrm{ind}}(x|\psi)$ may incorporate the effects of mate choice in the sexual case or intrinsic duplication rate in the asexual case, and $d_{\textrm{int}}(x|\psi)$ may model death due to intraspecific competition for resources. Substituting equation \eqref{BD_for_kimura} into \eqref{deterministic_traj}, we obtain
\label{PDE_for_kimura}
\frac{\partial \psi}{\partial t}(x,t) = w(x|\psi)\psi(x,t) + \mu b_{\textrm{mut}}(x|\psi)
where we have defined $w(x|\psi) \coloneqq b_{\textrm{int}}(x|\psi) - d_{\textrm{int}}(x|\psi)$, which can be thought of as the (Malthusian) `fitness' of the phenotype $x$. To track population numbers and trait frequencies, we follow the approach of \citep{week_white_2021} and define
\label{pop_size_and_freq_for_kimura}
N_K(t) &\coloneqq \int\limits_{\mathcal{T}}\psi(x,t)dx\\
p(x,t) &\coloneqq \frac{\psi(x,t)}{N_K(t)}
We can also define the population mean fitness as:
\label{mean_fitness_for_kimura}
\overline{w}(t) = \int\limits_{\mathcal{T}}w(x|\psi)p(x,t)dx
Using the chain rule in the definition of $p(x,t)$, we can calculate:
\frac{\partial p}{\partial t} &= \frac{1}{N_K(t)}\frac{\partial \psi}{\partial t}(x,t) - \frac{\psi(x,t)}{N^2_K(t)}\frac{d N_K}{dt}\\
&= \frac{1}{N_K(t)}\frac{\partial \psi}{\partial t}(x,t) - \frac{\psi(x,t)}{N^2_K(t)}\int\limits_{\mathcal{T}}\frac{\partial \psi}{\partial t}(y,t)dy
Where we have used the definition of $N_K(t)$ and assumed that integrals and derivatives commute in the second line. Substituting \eqref{PDE_for_kimura}, we now obtain
\frac{\partial p}{\partial t} &= \frac{1}{N_K(t)}\left[w(x|\psi)\psi(x,t) + \mu b_{\textrm{mut}}(x|\psi)\right] - \frac{\psi(x,t)}{N^2_K(t)}\int\limits_{\mathcal{T}}w(y|\psi)\psi(y,t) + \mu b_{\textrm{mut}}(y|\psi)dy\\
&= w(x|\psi)p(x,t) + \frac{\mu}{N_{K}(t)} b_{\textrm{mut}}(x|\psi) - p(x,t)\left(\int\limits_{\mathcal{T}}w(y|\psi)p(y,t)dy+\frac{\mu}{N_K(t)}\int\limits_{\mathcal{T}} b_{\textrm{mut}}(y|\psi)dy\right)
where we have used the definition of $p(x,t)$ in the second line. Using \eqref{mean_fitness_for_kimura} and rearranging the terms gives us:
\label{cts_replicator_mutator}
\setlength{\fboxsep}{2\fboxsep}\boxed{\frac{\partial p}{\partial t}(x,t) = \left[w(x|\psi) - \overline{w}(t)\right]p(x,t)+\frac{\mu}{N_K(t)}\left[b_{\textrm{mut}}(x|\psi) - p(x,t)\int\limits_{\mathcal{T}} b_{\textrm{mut}}(y|\psi)dy\right]}
This is a continuous version of the replicator-mutator equation when each $x$ is viewed as a strategy. It also yields Kimura's continuum-of-alleles model when each $x$ is viewed as an allele, $b_{\textrm{mut}}(x|\psi)$ takes the form of a convolution of $\psi(x,t)$ with a mutation kernel, and the trait space is the entire real line, \emph{i.e.} $\mathcal{T} = \mathbb{R}$. To see this, let $b_{\textrm{mut}}(y|\psi) = \int\limits_{\mathbb{R}}m(y-z)\psi(z,t)dz$, where $m(x)$ is a mutation kernel, which by definition is normalized such that $\int\limits_{\mathbb{R}}m(x)dx = 1$. Let us further note that we have implicitly been assuming that the total number of individuals (scaled by $K$) remains finite at all times, \emph{i.e.} $N_K(t) = \int\limits_{\mathbb
R}\psi(x,t)dx < \infty \ \forall \ t$. Thus, $m(x)\psi(y,t) \in  \mathcal{L}^{1}(\mathbb{R}\times\mathbb{R}) \ \forall \ t$ and we can use the Fubini-Tonnelli theorem to interchange the order of integration of iterated integrals of $m(y-z)\psi(y)$. We are now ready to evaluate the rightmost integral of \eqref{cts_replicator_mutator}.\\
We have: 
\int\limits_{\mathbb{R}} b_{\textrm{mut}}(y|\psi)dy &= \int\limits_{\mathbb{R}}\int\limits_{\mathbb{R}} m(y-z)\psi(z,t)dzdy\nonumber\\
&= \int\limits_{\mathbb{R}}\int\limits_{\mathbb{R}} m(y-z)\psi(z,t)dydz\nonumber\\
&= \int\limits_{\mathbb{R}}\psi(z,t)\left(\int\limits_{\mathbb{R}} m(y-z)dy\right)dz\nonumber\\
&= \int\limits_{\mathbb{R}}\psi(z,t)\int\limits_{\mathbb{R}} m(u)dudz\nonumber\\
&= \int\limits_{\mathbb{R}}\psi(z,t)dz\int\limits_{\mathbb{R}} m(u)du\nonumber\\
&= N_K(t)\int\limits_{\mathbb{R}} m(u)du\label{convolution_integral_intermediate}
where we have used the Fubini-Tonnelli theorem to go from the first step to the second, and have made the substitution $u = y-z$ to go from the third to the fourth step. We then note that since $m$ is a kernel, it satisfies $\int\limits_{\mathbb{R}} m(u)du = 1$, and \eqref{convolution_integral_intermediate} therefore becomes $\int\limits_{\mathbb{R}} b_{\textrm{mut}}(y|\psi)dy = N_K(t)$. Substituting this in \eqref{cts_replicator_mutator}, we have

>>>>

Raymond Pearl, one of the pioneers of mathematical ecology as a discipline, wrote in 1927 that ``What we want to know is how the biological forces of natality and mortality are so integrated and correlated in their action as to lead to a final result in size of population which follows this particular curve rather than some other one'' \citep{pearl_growth_1976}. Pearl realized 95 years ago that population dynamics must be ultimately explained by the mechanistic processes of birth and death. Today, there are mounting calls for more mechanistic models of evolution that are grounded in these fundamental processes of birth and death \citep{geritz_mathematical_2012,doebeli_towards_2017}. Ecologists also increasingly recognize that incorporating stochasticity is vital to developing more realistic ecological models \citep{hastings_transients_2004, coulson_skeletons_2004, boettiger_noise_2018, shoemaker_integrating_2020,schreiber_does_2022} and does more than `add noise' to deterministic expectations. Individual-based models, where ecological rules are specified at the level of the individual, are a powerful mathematical tool for mechanistic descriptions of stochastic population dynamics \citep{black_stochastic_2012}. Birth-death processes are a very general class of stochastic processes that can be used to capture a wide range of eco-evolutionary processes. `System-size expansions' and their subsequent analysis using the `weak noise approximation' are common tools for analyzing stochastic birth-death processes that are well-known in the statistical physics and applied mathematics communities \citep{gardiner_stochastic_2009}. However, these tools remain relatively underappreciated in ecology, despite being relatively easy to understand and extremely well-motivated in scenarios germane to ecology and evolution. Here, I present a formulation of population dynamics constructed from first principles grounded in birth-death processes. To facilitate readership by a broad audience, only passing familiarity with calculus (derivatives, integrals, Taylor expansions) and probability are assumed. Familiarity with stochastic calculus is helpful for some sections but is not required, and I provide a brief introduction below. I begin by introducing birth-death processes, SDEs, and the Fokker-Planck equation in a non-technical manner, and providing the intuition for system-size expansions and weak noise analysis in ecological systems. In section \ref{sec_1D_processes}, I show how fluctuating population size of populations of identical individuals can be tracked through a one-dimensional birth-death process. I introduce a description of the system via a `master equation', and then conduct a `system-size expansion' to obtain a Fokker-Planck equation for the system. Finally, I conduct a weak noise approximation to arrive at a linear Fokker-Planck equation which can be solved exactly using some stochastic calculus to arrive at a closed-form solution given by a time-dependent Ornstein-Uhlenbeck process. As an example, I illustrate the complete process for a stochastic version of the logistic equation. In section \ref{sec_nD_processes}, I present a multivariate process to describe the evolution of discretely varying traits, and, as before, illustrate the system size expansion and the weak noise approximation. Under mild assumptions, we show that the deterministic limit of this process is the well-known replicator-mutator equation (or equivalently, Eigen's quasispecies equation), thus establishing the microscopic basis of well-known equations in evolutionary game theory. I also show that the mean value of the trait in the population changes according to the Price equation in the deterministic limit. Chapter \ref{chap_infD_processes} introduces a function-valued process to model the evolution of quantitative traits such as body size, which can take on uncountably many values. This function-valued process can then also be analyzed via a system-size approximation to arrive at a `functional' Fokker-Planck equation, in terms of functional derivatives. Under mild assumptions, we show that classic equations such as Kimura's infinite alleles model and the canonical equation of adaptive dynamics can be derived as the deterministic limits of this stochastic process. We also conduct a weak noise approximation to arrive at a linear functional Fokker-Planck equation.  Sections \ref{sec_1D_processes} and \ref{sec_nD_processes} apply techniques that are well-known in physics, as applied to population biology. In contrast, chapter \ref{chap_infD_processes}, to the best of my knowledge, is entirely original. Chapter \ref{chap_infD_processes} also presents a simple heuristic derivation of quantitative genetics models and adaptive dynamics from stochastic first principles that is much simpler than the rigorous mathematical derivations grounded in measure theory and martingale/markov theory that are currently standard reference in theoretical ecology \citep{champagnat_individual_2008}. I illustrate the utility of all this abstract formalism through examples in \ref{chap_examples}.

@@

\section{Warm up: One-dimensional processes for population size}\label{sec_1D_processes}
The simplest birth-death processes are those in which the state at any time can be characterized by a single number. Populations of identical individuals are an obvious example of such a system. The mathematics below are adapted from sections 6.3 and 7.2 of \citep{gardiner_stochastic_2009}.

@@
